---
title: "Modelsummary package - datasummary()"
author: "Robert Cline"
date: "11/15/2021"
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) 

library(tidyverse)
library(modelsummary)  
library(palmerpenguins)
library(data.table)
library(here)
## https://github.com/allisonhorst/palmerpenguins  
## collected at the Palmer Station in Antarctica by Gorman, Williams & Fraser (2014)
```  

Tutorial for [Vincent Arel Bundock's *modelsummary* package](https://vincentarelbundock.github.io/modelsummary/index.html)

[model plot](https://vincentarelbundock.github.io/modelsummary/articles/modelplot.html)

```{r}
library(data.table)

owid0 <- fread(here("data", "owid-covid-data.csv"))
    # select(people_vaccinated_per_hundred, people_fully_vaccinated_per_hundred) %>% 
owid <- owid0 %>% 
  mutate(all_vax = people_vaccinated_per_hundred,
         fully_vax = people_fully_vaccinated_per_hundred) %>% 
  select(location, date,
         all_vax = people_vaccinated_per_hundred,
         fully_vax = people_fully_vaccinated_per_hundred) %>% 
  filter((location == "Australia" | location=="Brazil" | location == "Canada" | location  == "China" | location == "France" | location == "Germany" | location == "India" | location == "Italy" | location == "Mexico" | location == "South Africa" | location == "South Korea" | location == "Spain" | location == "United Kingdom" | location == "United States") & date == "2021-10-28") %>% 
  drop_na(fully_vax)  

  
 datasummary_skim(owid)

```

```{r}
 datasummary_skim(owid, type="categorical")

```
```{r}
datasummary_skim(penguins, type = "categorical")
```


### Crosstabs, frequencies, correlations, balance  
(a.k.a. “table 1”), and more  
* Many output formats: HTML, LaTeX, Microsoft Word and Powerpoint, Text/Markdown, PDF, RTF, or Image files.  
* Embed tables in Rmarkdown or knitr dynamic documents.  
* Customize the appearance of tables with the gt, kableExtra or flextable packages. The possibilities are endless!  
* This tutorial will show how to draw tables like these (and more!):

[datasummary built around tables package](https://vincentarelbundock.github.io/modelsummary/articles/datasummary.html)  

In addition to datasummary, the modelsummary package includes a “family” of companion functions named datasummary_*. These functions facilitate the production of standard, commonly used tables. This family currently includes:  

* *datasummary()*: Flexible function to create custom tables using 2-sided formulae.  
* *datasummary_balance()*: Group characteristics (e.g., control vs. treatment)  
* *datasummary_correlation()*: Table of correlations.  
* *datasummary_skim()*: Quick summary of a dataset.  
* *datasummary_df()*: Create a table from any dataframe.  
* *datasummary_crosstab()*: Cross tabulations of categorical variables.  

In the next three sections, we illustrate how to use datasummary_balance, datasummary_correlation, datasummary_skim, and datasummary_crosstab. Then, we dive into datasummary itself to highlight its ease and flexibility.  

Customize the appearance of tables with the *gt, kableExtra or flextabl*e* packages. The possibilities are endless!

datasummary_skim(penguins)  

datasummary  


### datasummary tables are specified using a *2-sided formula, divided by a tilde ~.*   
* The left-hand side describes the rows;  
* the right-hand side describes the columns.   

To illustrate how this works, we will again be using the palmerpenguins dataset:

To display the flipper_length_mm variable as a row and the mean as a column, we type:

```{r echo=TRUE}
datasummary(flipper_length_mm ~ Mean,
            data = penguins)
```

To flip rows and columns, we flip the left and right-hand sides of the formula:  

```{r echo=TRUE}
datasummary(Mean ~ flipper_length_mm,
            data = penguins)
```

### Custom summary functions  
* The Mean function is a shortcut supplied by modelsummary; and  
* it is equivalent to mean(x,na.rm=TRUE).   

Since the flipper_length_mm variable includes missing observation, using the mean formula (with default na.rm=FALSE) would produce a missing/empty cell:

```{r}
datasummary(flipper_length_mm ~ mean,
            data = penguins)
```

***modelsummary*** supplies these functions:   
Mean, SD, Min, Max, Median, P0, P25, P50, P75, P100, Histogram, and a few more (see the package documentation).  

Users are also free to create and use their own custom summaries. Any R function which takes a vector and produces a single value is acceptable.   

For example, the Range functions return a numerical value, and the MinMax returns a string:  


Example1
```{r echo=TRUE}
Range <- function(x) max(x, na.rm = TRUE) - min(x, na.rm = TRUE)

datasummary(flipper_length_mm ~ Range,
            data = penguins)
```

Example2  
```{r echo=TRUE}
MinMax <- function(x) paste0('[', min(x, na.rm = TRUE), ', ', max(x, na.rm = TRUE), ']')

datasummary(flipper_length_mm ~ MinMax,
            data = penguins)
```


### Concatenating with +  
To include more rows and columns, we use the + sign:

```{r}
datasummary(flipper_length_mm + body_mass_g ~ Mean + SD,
            data = penguins)
```

Sometimes, it can be cumbersome to list all variables separated by + signs. The **All() function** is a useful shortcut:

```{r echo=TRUE}
datasummary(All(penguins) ~ Mean + SD,
            data = penguins)
```
By default, All selects all numeric variables. This behavior can be changed by modifying the function’s arguments. See ?All for details  

### Nesting with *  

datasummary can nest variables and statistics inside categorical variables using the * symbol. When applying the the * operator to factor, character, or logical variables, columns or rows will automatically be nested.   

For instance, if we want to display separate means for each value of the variable sex, we use mean * sex:  

```{r echo=TRUE}
datasummary(flipper_length_mm + body_mass_g ~ mean * sex,
            data = penguins)
```

We can use parentheses to nest several terms inside one another, using a call of this form: x * (y + z). Here is an example with nested columns:  


### Nesting terms inside of one another:  
```{r}
datasummary(body_mass_g ~ sex * (mean + sd),
            data = penguins)
```

### Nested rows:  
```{r}

datasummary(sex * (body_mass_g + flipper_length_mm) ~ mean + sd,
            data = penguins)
```

**Order terms**:  The order in which terms enter the formula determines the order in which labels are displayed. For example, this shows island above sex:  

```{r}
datasummary(flipper_length_mm + body_mass_g ~ mean * island * sex,
            data = penguins)

```

This shows sex above island values:  
```{r}
datasummary(flipper_length_mm + body_mass_g ~ mean * sex * island,
            data = penguins)
```

By default, datasummary omits column headers with a single value/label across all columns, except for the header that sits just above the data. If the header we want to see is not displayed, we may want to reorder the terms of the formula. **To show all headers**, set *sparse_header=FALSE*:

```{r}
datasummary(flipper_length_mm + body_mass_g ~ mean * sex * island,
            data = penguins,
            sparse_header = FALSE)
```

When using sparse_header=FALSE, it is often useful to insert Heading() * in the table formula, in order to rename or omit some of the labels manually. Type ?tables::Heading for details and examples.

### Renaming with = 
Personally, I prefer to rename variables and values before drawing my tables, using backticks when variable names include whitespace. For example,

```{r echo=TRUE}
tmp <- penguins %>%
       select(`Flipper length (mm)` = flipper_length_mm,
              `Body mass (g)` = body_mass_g)

datasummary(`Flipper length (mm)` + `Body mass (g)` ~ Mean + SD,
            data = tmp)
```

However, thanks to the tables package, datasummary offers two additional mechanisms to rename. First, we can wrap a term in parentheses and use the equal = sign: (NewName=OldName):

```{r}
datasummary((`Flipper length (mm)` = flipper_length_mm) + (`Body mass (g)` = body_mass_g) ~
            island * ((Avg. = Mean) + (Std.Dev. = SD)),
            data = penguins)
```

### Heading() function:  

```{r echo=TRUE}
datasummary(Heading("Flipper length (mm)") * flipper_length_mm + Heading("Body mass (g)") * body_mass_g ~ island * (Mean + SD),
            data = penguins)
```

The Heading function also has a ***nearData argument*** which can be useful in cases where some rows are nested but others are not. Compare the last row of these two tables:  

```{r echo=TRUE}
datasummary(sex * (flipper_length_mm + bill_length_mm) + Heading("Body mass (g)") * body_mass_g ~ Mean + SD,
            data = penguins)
```

```{r echo=TRUE}
datasummary(sex * (flipper_length_mm + bill_length_mm) + Heading("Body mass (g)", nearData=FALSE) * body_mass_g ~ Mean + SD,
            data = penguins)
```

### Counts and Percentages -> Frequency Table  
The tables package allows datasummary to use neat tricks to produce frequency tables:  

* Add a N to the right-hand side of the equation.  
* Add Percent() to the right-hand side to calculate the percentage of observations in each cell.  
* Add 1 to the left-hand side to include a row with the total number of observations:
```{r echo=TRUE}
datasummary(species * sex + 1 ~ N + Percent(),
            data = penguins)
```
Note that the Percent() function accepts a denom argument to determine if percentages should be calculated row or column-wise, or if they should take into account all cells.  

### Factor
The * nesting operator that we used above works automatically for factor, character, and logical variables. Sometimes, it is convenient to use it with other types of variables, such as binary numeric variables. In that case, we can wrap the variable name inside a call to Factor(). This allows us to treat a variable as factor, without having to modify it in the original data. For example, in the mtcars data, there is a binary numeric variable call am. We nest statistics within categories of am by typing:

```{r echo=TRUE}
datasummary(mpg + hp ~ Factor(am) * (mean + sd),
            data = mtcars)
```

### Arguments: na.rm=TRUE  
We can pass any argument to the summary function by including a call to **Arguments()**. For instance, there are missing values in the flipper_length_mm variable of the penguins dataset. Therefore, the standard mean function will produce no result, because its default argument is na.rm=FALSE. We can change that by calling:  

```{r echo=TRUE}
datasummary(flipper_length_mm ~ mean + mean*Arguments(na.rm=TRUE),
            data = penguins)
```

Notice that there is an empty cell (NA) under the normal mean function, but that the mean call with Arguments(na.rm=TRUE) produced a numeric result.  

We can pass the same arguments to multiple functions using the parentheses:  

```{r}
datasummary(flipper_length_mm ~ (mean + sd) * Arguments(na.rm=TRUE),
            data = penguins)
```

**Note** that the shortcut functions that modelsummary supplies use na.rm=TRUE by default, so we can use them directly without arguments, even when there are missing values:

```{r}
datasummary(flipper_length_mm ~ Mean + Var + P75 + Range,
            data = penguins)
```

### Empty cells  
Sometimes, if we nest too much and the dataset is not large/diverse enough, we end up with empty cells. In that case, we add **DropEmpty()** to the formula:  

Without *DropEmpty()*  
```{r}
datasummary(island * species * body_mass_g ~ sex * (Mean + SD),
            data = penguins)
```

With *DropEmpty()*  
```{r}
datasummary(island * species * body_mass_g ~ sex * (Mean + SD) * DropEmpty(),
            data = penguins)
```

### Logical subsets  
Cool stuff is possible with logical subsets:  

```{r}
datasummary((bill_length_mm > 44.5) + (bill_length_mm <= 44.5) ~ Mean * flipper_length_mm,
            data = penguins) 
```
See the tables package documentation for more details and examples.  

### output  
All functions in the datasummary_* family accept the same output argument. Tables can be saved to several file formats:  

```{r}
f <- flipper_length_mm ~ island * (mean + sd)
datasummary(f, data = penguins, output = 'table.html')
datasummary(f, data = penguins, output = 'table.tex')
datasummary(f, data = penguins, output = 'table.docx')
datasummary(f, data = penguins, output = 'table.pptx')
datasummary(f, data = penguins, output = 'table.md')
datasummary(f, data = penguins, output = 'table.rtf')
datasummary(f, data = penguins, output = 'table.jpg')
datasummary(f, data = penguins, output = 'table.png')
```

They can be returned in human-readable data.frames, markdown, html, or LaTeX code to the console:  

```{r}
datasummary(f, data = penguins, output = 'data.frame')
datasummary(f, data = penguins, output = 'markdown')
datasummary(f, data = penguins, output = 'html')
datasummary(f, data = penguins, output = 'latex')  
```

datasummary can return objects compatible with the gt, kableExtra, flextable, and huxtable table creation and customization packages:

```{r}
datasummary(f, data = penguins, output = 'gt')
datasummary(f, data = penguins, output = 'kableExtra')
datasummary(f, data = penguins, output = 'flextable')
datasummary(f, data = penguins, output = 'huxtable')
```

Please note that hierarchical or “nested” column labels are only available for these output formats: kableExtra, gt, html, rtf, and LaTeX. When saving tables to other formats, nested labels will be combined to a “flat” header.

### fmt  
The fmt argument allows us to set the printing format of numeric values. It accepts a single number representing the number of digits after the period, or a string to be passed to the sprintf function. For instance, setting fmt="%.4f" will keep 4 digits after the dot (see ?sprintf for more options):  
```{r}


datasummary(flipper_length_mm ~ Mean + SD,
            fmt = 4,
            data = penguins)
```
*requires library(huxtable)*  WHICH IS INSTALLED  

If we do not want datasummary to format numbers, and we want to keep the numerical values instead of formatted strings, set *fmt=NULL*.  

This can be useful when post-processing tables with packages like gt, which allow us to transform cells based on their numerical content (this gt table will be omitted from PDF output):

```{r}
library(gt)

datasummary(All(mtcars) ~ Mean + SD,
            data = mtcars,
            fmt = NULL,
            output = 'gt') %>%
    tab_style(style = cell_fill(color = "pink"),
              locations = cells_body(rows = Mean > 10, columns = 2))

```

### title, notes  
datasummary includes the same title and notes arguments as in modelsummary:  

```{r}
datasummary(All(penguins) ~ Mean + SD,
            data = penguins,
            title = 'Statistics about the famous Palmer Penguins.',
            notes = c('A note at the bottom of the table.'))
```

### align  
We can align columns using the align argument. *Align* should be a string of length equal to the number of columns, and which includes only the letters “l”, “c”, or “r”:  

```{r}
datasummary(flipper_length_mm + bill_length_mm ~ Mean + SD + Range,
            data = penguins,
            align = 'lrcl')
```

### add_rows  
```{r}
new_rows <- data.frame('Does',
                       2,
                       'plus',
                       2,
                       'equals',
                       5,
                       '?')
datasummary(flipper_length_mm + body_mass_g ~ species * (Mean + SD),
            data = penguins,
            add_rows = new_rows)
```

### add_columns  
```{r}
new_cols <- data.frame('New Stat' = runif(2))
datasummary(flipper_length_mm + body_mass_g ~ species * (Mean + SD),
            data = penguins,
            add_columns = new_cols)
```

### Histograms  
The datasummary family of functions allow users to display *in-line spark-style histograms* to describe the distribution of the variables. For example, the datasummary_skim produces such a histogram:

```{r}
tmp <- mtcars[, c("mpg", "hp")]
datasummary_skim(tmp)
```

### Histograms with different output backend.  

Each of the histograms in the table above is actually an SVG image, produced by the kableExtra package. For this reason, the histogram will not appear when users use a different output backend, such as gt, flextable, or huxtable.

The datasummary function is incredibly flexible, but it does not include a histogram option by default. 

**Customize datasummary output:**  
Here is a simple example of how one can customize the output of datasummary. We proceed in 4 steps:

1. Normalize the variables and store them in a list  
2. Create the table with datasummary, making sure to include 2 “empty” columns. In the example, we use a simple function called *emptycol* to fill those columns with empty strings.  
3. Add the histograms or boxplots using functions from the *kableExtra* package.  
```{r}
library(kableExtra)

tmp <- mtcars[, c("mpg", "hp")]

# create a list with individual variables
# remove missing and rescale
tmp_list <- lapply(tmp, na.omit)
tmp_list <- lapply(tmp_list, scale)

# create a table with `datasummary`
# add a histogram with column_spec and spec_hist
# add a boxplot with colun_spec and spec_box
emptycol = function(x) " "
datasummary(mpg + hp ~ Mean + SD + Heading("Boxplot") * emptycol + Heading("Histogram") * emptycol, data = tmp) %>%
    column_spec(column = 4, image = spec_boxplot(tmp_list)) %>%
    column_spec(column = 5, image = spec_hist(tmp_list))
```

If you want a simpler solution, you can try the Histogram function which works in datasummary automatically and comes bundled with modelsummary.  

The downside of this function is that it uses Unicode characters to create the histogram. *This kind of histogram may not display well with certain typefaces or on some operating systems (Windows!).  
**Does not display histogram in Windows** 
```{r}
datasummary(mpg + hp ~ Mean + SD + Histogram, data = tmp)
```

### Missing values  
At least 3 distinct issues can arise related to missing values.

### Functions and na.rm  
An empty cell can appear in the table when a statistical function returns **NA** instead of a numeric value. In those cases, you should:  

1. Make sure that your statistical function (e.g., mean or sd) uses na.rm=TRUE by default  
2. **Use the Arguments strategy** to *set na.rm=TRUE* (see the Arguments section of this vignette).  
3. Use a convenience function supplied by modelsummary, where na.rm is TRUE by default: Mean, SD, P25, etc.  

### Empty crosstab cells  
An empty cell can appear in the table when a crosstab is deeply nested, and there are no observations for a given combination of covariates. In those cases, you can use the * **DropEmpty pseudo-function**. See the “*Empty cells*” section of this vignette for examples.

### Factors  

By default, the factor function in R does not assign a distinct factor level to missing values: the factor function’s exclude argument is set to NA by default. To ensure that NAs appear in your table, make sure you set exclude=NULL when you create the factor.

Internally, the *datasummary_balance* and *datasummary_crosstab* functions convert logical and character variables to factor with the *exclude=NULL argument*. This means that **NAs** will appear in the table as distinct rows/columns.   

**If you do not want NAs to appear in your table**, convert them to factors yourself ahead of time. For example:

```{r}
mycars <- mtcars[, c("cyl", "mpg", "hp", "vs")]
mycars$cyl[c(1, 3, 6, 8)] <- NA
mycars$cyl_nona <- factor(mycars$cyl)
mycars$cyl_na <- factor(mycars$cyl, exclude = NULL)

datasummary_crosstab(cyl_nona ~ vs, data = mycars)
```

```{r}
datasummary_crosstab(cyl_na ~ vs, data = mycars)
```


### Appearance  - Customizing tables
In the Appearance Vignette we saw how to customize the tables produced by the *modelsummary function*.   \

The same customization possibilities are also available for all functions in the *datasummary_* family of functions*.   

Indeed, we can *customize* the tables produced by datasummary using the functions provided by *gt, kableExtra, flextable, and huxtable*. For instance, to customize tables using kableExtra, we can do:

This chunk does not run.  

* Error in xml_children(x)[[search]] : subscript out of bounds  
* [Issue reported 2021-11-16](https://github.com/vincentarelbundock/modelsummary/issues/414)
```{r eval=FALSE}
library(kableExtra)
library(modelsummary)
library(palmerpenguins)

datasummary(All(penguins) ~ sex * (Mean + SD),
            data = penguins,
            output = 'kableExtra') %>%
  row_spec(3, background = 'cyan', color = 'red')
```


### Example of customizing a table with *gt* package.  
```{r}
library(gt)

adelie <- function(x) web_image('https://user-images.githubusercontent.com/987057/85402702-20b1d280-b52a-11ea-9950-f3a03133fd45.png', height = 100)
gentoo <- function(x) web_image('https://user-images.githubusercontent.com/987057/85402718-27404a00-b52a-11ea-9ad3-dd7562f6438d.png', height = 100)
chinstrap <- function(x) web_image('https://user-images.githubusercontent.com/987057/85402708-23acc300-b52a-11ea-9a77-de360a0d1f7d.png', height = 100)

cap <- 'Flipper lengths (mm) of the famous penguins of Palmer Station, Antarctica.'
f <- (`Species` = species) ~ (` ` = flipper_length_mm) * (`Distribution` = Histogram) + flipper_length_mm * sex * ((`Avg.` = Mean)*Arguments(fmt='%.0f') + (`Std. Dev.` = SD)*Arguments(fmt='%.1f'))
datasummary(f,
            data = penguins,
            output = 'gt',
            title = cap,
            notes = 'Artwork by @allison_horst',
            sparse_header = TRUE) %>%
    text_transform(locations = cells_body(columns = 1, rows = 1), fn = adelie) %>%
    text_transform(locations = cells_body(columns = 1, rows = 2), fn = chinstrap) %>%
    text_transform(locations = cells_body(columns = 1, rows = 3), fn = gentoo) %>%
    tab_style(style = list(cell_text(color = "#FF6700", size = 'x-large')), locations = cells_body(rows = 1)) %>%
    tab_style(style = list(cell_text(color = "#CD51D1", size = 'x-large')), locations = cells_body(rows = 2)) %>%
    tab_style(style = list(cell_text(color = "#007377", size = 'x-large')), locations = cells_body(rows = 3)) %>%
    tab_options(table_body.hlines.width = 0, table.border.top.width = 0, table.border.bottom.width = 0) %>%
    cols_align('center', columns = 3:6)
```

[Reference](https://vincentarelbundock.github.io/modelsummary/reference/index.html)  

