---
title: "Frequency Tables"
author: "ehsanx"
date: "8/24/2021"
output: html_document
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)  


library(tidyverse)
library(palmerpenguins)
library(summarytools)
library(janitor)
library(gt)
library(here)
```
### Summarytools package for frequency distributions.
* Missing row names
* Computes data as an array.
* Must convert to a data.frame with data.table package to pring.

```{r}

pengs <- freq(penguins$species, order = "freq")

dt1 <- data.table::as.data.table(pengs)
```
 
 
### Print data.table with gt()  
* eval = FALSE   will not knit.  
 
```{r eval=FALSE}

options(scipen=999)
# library(gt)
# pengs <- tibble(pengs)
gt(dt1)
``` 

### Frequency Table with two variables
```{r}

mtcars %>%
  group_by(am, gear) %>%
  dplyr::summarise(n = n()) %>%
  mutate(freq = n / sum(n)) %>% 
  ungroup()  ## Ungroup for later manipulatons
```
### Relative Frequency Table 
```{r}
data(mtcars)

mtcars %>%
  group_by (am, gear) %>%
  dplyr::summarise (n=n()) %>%
  mutate(rel.freq = paste0(round(100 * n/sum(n), 0), "%")) %>% 
  ungroup()

```


### Relative Frequency table with custom function
```{r}
as.rel_freq <- function(x, rel_freq_col = "rel.freq", ...) {
    class(x) <- c("rel_freq", class(x))
    attributes(x)[["rel_freq_col"]] <- rel_freq_col
    x
}

print.rel_freq <- function(x, ...) {
    freq_col <- attributes(x)[["rel_freq_col"]]
    x[[freq_col]] <- paste0(round(100 * x[[freq_col]], 0), "%")   
    class(x) <- class(x)[!class(x)%in% "rel_freq"]
    print(x)
}

mtcars %>%
  group_by (am, gear) %>%
  dplyr::summarise (n=n()) %>%
  mutate(rel.freq = n/sum(n)) %>%
  as.rel_freq() %>% 
  ungroup()
``` 


### Using scipen
* Positive values bias towards fixed and negative towards scientific notation
* turn off options(scipen = 999)
* turn on  options(scipen = 0)
* to show scientific notation:
  - options(scipen = -100)
```{r}

# Turn scipen off with 
getOption("scipen") 
options("scipen"= -100) 
mtcars %>%
dplyr::count(am, gear) %>% 
mutate(freq = (n / sum(n)) * 100)
```

```{r}
mtcars %>% 
  dplyr::count(am, gear) %>% 
  add_count(am, wt = n) %>% 
  mutate(percent = n / nn)
```


```{r}

mtcars %>% 
dplyr::count(am, gear)
```

```{r}
mtcars %>%
  group_by(am, gear) %>%
  dplyr::summarise(n = n()) %>%
  mutate(freq = n / sum(n)) %>% 
  mutate(cumsum = cumsum(n)) %>% 
  mutate(cumsum_freq = cumsum(freq))
```


### Percentage using adorn formatting 
* Set eval = FALSE.  Will not knit.  Cannot find function *"adorn_ns"*
```{r eval=FALSE}
mtcars %>% 
  dplyr::count(cyl, gear, name = "proportion") %>% 
  adorn_percentages("col", na.rm = TRUE, proportion) %>% 
  adorn_totals(where = "row", fill = "-", na.rm = TRUE, 
               name = "Total", proportion) %>% 
  adorn_pct_formatting(digits = 1, rounding = "half up", 
                       affix_sign = TRUE, proportion) %>% 
  adorn_ns(position = "rear", ns = attr(., "core"), proportion)
``` 


### Frequency, CumFrequency and RelFrequency  
* Using *table()*  and  *cbind()*  
```{r}

x <- table(mpg$model)
cbind( Freq=table(x), Cumul=cumsum(table(x)), RelFreq=prop.table(table(x)))
 
```

### Three ways to get frequecy tables  

```{r}
library(readr)
nhanes <- read_csv(here("data", "nhanes.csv"))
View(nhanes)

``` 



### prop.table()
```{r}
nhanes %>% 
  clean_names() %>% 
  group_by(education) %>% 
  dplyr::summarize(n=n()) %>% 
  mutate(percent = round(prop.table(n), 2)*100) %>% 
  mutate(prop = round(prop.table(n), 2))
  
```

### Count 
```{r eval=FALSE}
nhanes %>% 
   dplyr::mutate(Edcount = count(Education))

%>% 
  dplyr::mutate(percent = prop.table(n)) %>% 
  dplyr::mutate(rel.freq = paste0(round(100 * n/sum(n), 0), "%"))
``` 

```{r eval=FALSE}
nhanes %>% 
  clean_names() %>% 
  filter(!is.na(education)) %>% 
  mutate(n=count(education))
```



### tally() and cumsum()
* tally() is short-hand for summarise() 

```{r}
nhanes %>%
  group_by(Education) %>%
  tally() %>% 
  mutate(cumulative = cumsum(n))

```

### count() is a short-hand for group_by() + tally()  
```{r}
nhanes %>% 
  dplyr::count(Education)

# count(nhanes$Education)

```

### Note that if the data is already grouped, count() adds
# an additional group that is removed afterwards
mtcars %>% group_by(gear) %>% count(carb)

```{r} 
nhanes %>%
  dplyr::count(Education, Gender)

```




### tably package from janitor package  
* valid percent is the proportion without counting the NAs  
```{r}

nhanes %>% 
  clean_names() %>% 
  tabyl(education)
```

