---
title: "Cumulative Freq Distribution of Old Faithful"
author: "R Tutorial"
date: "8/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[R Tutorial Cumulative Freq Distr](http://www.r-tutor.com/elementary-statistics/quantitative-data/cumulative-frequency-distribution)  

### Data Built-in data(faithful) 
* eruptions = the duration of the geyser eruptions
* waiting = length of the waiting period
* there is a correlation between the two variables shown in the scatter plot

```{r}
duration = faithful$eruptions      # the eruption durations 
waiting = faithful$waiting         # the waiting interval 
plot(duration, waiting,           # plot the variables 
   xlab="Eruption duration (seconds)",        # x−axis label 
   ylab="Time waited (minutes")              # y−axis label
abline(lm(waiting ~ duration))
```

```{r}
data(faithful)
```

### Duration of eruptions
The range of duration is from `r min(duration)` seconds to `r max(duration)` seconds 


```{r}
range(duration)
```
### Eruption intervals  
Break the range into non-overlapping sub-intervals by defining a sequence of equal distance break points. If we round the endpoints of the interval [1.6, 5.1] to the closest half-integers, we come up with the interval [1.5, 5.5]. Hence we set the break points to be the half-integer sequence { 1.5, 2.0, 2.5, ... }.  

```{r}
breaks = seq(1.5, 5.5, by=0.5)    # half-integer sequence 
breaks 
```

Classify the eruption durations according to the half-unit-length sub-intervals with cut.  
As the intervals are to be closed on the left, and open on the right, we set the right argument as FALSE.  

```{r}
duration.cut = cut(duration, breaks, right=FALSE)

# x <- table(mpg$model)
x <- duration.cut

cbind(Freq=table(x), Cumul=cumsum(table(x)), RelFreq=prop.table(table(x)))



```


Compute the frequency of eruptions in each sub-interval with the table function.  
```{r}
duration.freq <- table(duration.cut)
```
 
 The frequency distribution of the eruption duration is:  
```{r}
duration.freq
# duration.cut

```
 
 Use the cbind function to print the result in column format  
```{r}

cbind(duration.freq)
duration.freq
```
Per R documentation, you are advised to use the hist function to find the frequency distribution for performance reasons.

Exercise
Find the frequency distribution of the eruption waiting periods in faithful.  
```{r}
hist(duration,
     right=FALSE,
     ylim = c(0,80))
```


Find programmatically the duration sub-interval that has the most eruptions.  
```{r}
hist(duration,
     right=FALSE,
     ylim = c(0,80))

```

