---
title: "Univariate Analysis"
author: "Robert Cline"
date: "11/20/2021"
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Jaccard Similarity Test  


[jaccur similarity test](https://www.youtube.com/watch?v=mWL89jrWtj0)

[Jaccard similarity in R from r-bloggers](https://www.r-bloggers.com/2021/11/how-to-calculate-jaccard-similarity-in-r/)  

[Mathematics in R Markdown](https://rpruim.github.io/s341/S19/from-class/MathinRmd.html)

$Jaccard(A,B) =  \frac {|A \cup B|}{|A \cap B|}$  
 

Jaccard Example:  

```{r}  


a <- c(1,5,8,10,22,14,15,16,2,7)
b <- c(10,12,13,2,7,9,2,7,23,15)
```

Jaccard function 
```{r}
jaccard <- function(a, b) {
    intersection = length(intersect(a, b))
    union = length(a) + length(b) - intersection
    return (intersection/union)
}
```


The Jaccard Similarity between the two lists is 0.25. As mentioned above greater the number closer to the data sets.   
This also works with strings.  

* 0 = no similarity  
* 1 = match exactly  
```{r}
jaccard(a,b)
```


Calculate the distance between two sets  
* 1-jaccard(a, b)  

```{r}
a <- c(1,5,8,10,22,14,15,16,2,7)
b <- c(10,12,13,2,7,9,2,7,23,15)
1-jaccard(a, b)
```

If you’re looking for a way to calculate the Jaccard similarity matrix, the vegan package is a good place to start. Many other similarities/dissimilarity measures can be calculated with the vegdist() function.  

### [More tests from Statistics Globe](https://statisticsglobe.com/compare-vectors-and-find-differences-in-r)  


Identical ?  
* identical(a,b)  
* a=b  
```{r}

identical(a, b) 
```

Which vector elements are the same  
* a==b
```{r}

a==b
```


Which elements of first exist in the second using *%in%*  

```{r}
a %in% b
```

Find elements that exist in the first & second using *intersect()*  
```{r}

intersect(a, b)
```

Find elements that exist in the first, but not in the second  *setdiff()*  
```{r}

setdiff(a,b)
```

